---
title: "09_fruitTrees"
author: "zagor"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_caption: yes
    self_contained: yes
    fig_width: 9
    fig_height: 6
    toc: true
    toc_float:
      toc_collapsed: false
    toc_depth: 5
    number_sections: yes
    theme: flatly
    highlight: tango
  pdf_document:
    toc: yes
    toc_depth: '2'
  word_document:
    toc: yes
    toc_depth: '2'
editor_options:
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(dev = c('png', 'pdf'),  # this embeds pdf and crates scrolable blocks if pdf is first
                      fig.path = "../reports/fruitTrees/", # save each figure here
                      # dev = c('png'), 
                      fig.align = 'center', 
                      dpi = 300,
                      fig.height = 6, 
                      fig.width = 9 ,
                      # warning = FALSE, message = FALSE
                      echo = TRUE, message = TRUE, warning = TRUE
                      )
```



```{r,  echo=TRUE, warning=FALSE, message=FALSE}

rm(list = ls(all = TRUE))
gc()


library(magrittr)
library(data.table)
library(knitr)
library(ggplot2)
library(ComplexUpset)


`%!in%` = Negate(`%in%`)
`%nin%` = Negate(`%in%`)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))




```

# Ath gmm


```{r, Ath-gmm}

fp = file.path('..', 'intermediate')
fn = 'mercator_cleaned.txt.zip'
gmm = data.table::fread(file.path(fp, fn), header = TRUE, fill = TRUE)
ath.gmm = gmm[gmm$plant == 'ath', ]
setDT(ath.gmm)
ath.gmm[, plant := NULL]
ath.gmm[, source := NULL]
colnames(ath.gmm)[2:4] = paste('ath', colnames(ath.gmm)[2:4], sep = '_')

```


# Ath SKM & annotation

note: some duplicated ids in PSS


```{r, Ath-SKM-annot}

fp = file.path('..', 'input', 'ath-annot', 'Phytozome', 'PhytozomeV12', 
               'early_release', 'Athaliana_447_Araport11', 'annotation')
# fn = 'Araport11_GFF3_genes_transposons.current_utf8_attributes_CB.tsv'
fn = 'Athaliana_447_Araport11.geneName.txt'
gn = data.table::fread(file.path(fp, fn), header = FALSE, fill = TRUE)
colnames(gn)[2] = 'athName'
gn$V1 = sub('\\..*', '', gn$V1)
gn = gn[!duplicated(gn), ]


fn = 'Athaliana_447_Araport11.synonym.txt'
sn = data.table::fread(file.path(fp, fn), header = FALSE, fill = TRUE)
sn[, merged_column := apply(.SD, 1, function(x) {
  # Remove NA and empty strings
  x = x[!is.na(x) & x != ""]
  paste(x, collapse = " | ")
}), .SDcols = 2:ncol(sn)]
# Optionally, remove the original columns V2 to V15
sn[, (2:(ncol(sn)-1)) := NULL]
colnames(sn)[2] = 'athSynonims'
sn$V1 = sub('\\..*', '', sn$V1)
sn = sn[!duplicated(sn), ]


fp = file.path('..', 'input', 'SKM_2025-07-08')
fn = 'rxn-nodes-public.tsv'
pss = data.table::fread(file.path(fp, fn), header = TRUE, fill = TRUE)
ind = grep('^name$|^all_pathways|^short_name$', colnames(pss), value = TRUE)
pss = pss[, ..ind]
ind = grep('\\[', pss$name)
pss = pss[ind, ]

pss[, ids_string := stringr::str_extract(name, "(?<=\\[)[^\\]]+(?=\\])")]
pss[, ids_list := strsplit(ids_string, split = ",")]
max_ids = max(lengths(pss$ids_list))
for (i in seq_len(max_ids)) {
  pss[[paste0("id_", i)]] = sapply(pss$ids_list, function(x) ifelse(length(x) >= i, x[i], NA))
}
pss[, c("ids_string", "ids_list") := NULL]

pss_long = melt(
  pss,
  id.vars = c("name", "all_pathways", 'short_name'),       # Columns to keep as is
  measure.vars = patterns("^id_"),           # Columns to melt (all starting with "id_")
  variable.name = "id_num",                   # Name for the melted variable column
  value.name = "id"                           # Name for the melted value column
)

pss_long = pss_long[!is.na(id) & id != ""]
pss_long[, id_num := NULL]
pss_long[, name := NULL]
pss_long$id = sub('\\..*', '', pss_long$id)
pss_long = pss_long[!duplicated(pss_long), ]
table(duplicated(pss_long$id))

pss_long %>%
  dplyr::filter(id %in% id[duplicated(id)] & stringr::str_starts(id, "^AT")) %>%
  dplyr::arrange(id) %>%
  print()


pss_long = pss_long %>%
  dplyr::filter(stringr::str_starts(id, "AT")) %>%
  dplyr::group_by(id) %>%
  dplyr::summarise(
    dplyr::across(
      .cols = dplyr::everything(),
      .fns = ~ {
        vals = unique(na.omit(.))
        if (length(vals) > 1) paste(vals, collapse = " | ")
        else if (length(vals) == 1) vals
        else NA_character_
      }
    ),
    .groups = "drop"
  )

pss_long[pss_long == ""] = "-"
gn[is.na(gn)] = "-"
sn[is.na(sn)] = "-"

ath.annot = pss_long %>%
  dplyr::full_join(gn, by = c("id" = "V1")) %>%
  dplyr::full_join(sn, by = c("id" = "V1"))



```



**Note**: 35.2 bin matcheswill be ignored


# Abbreviations

| Plant Name                    | Label         | JCVI-MCScan | Compara Plants | Plaza | OrthoDB | FastOma | RBH  | Mercator |
|-------------------------------|---------------|-------------|----------------|-------|---------|---------|------|----------|
| Malus domestica               | apple         | *mdo_GDv1*  | *malus_domestica_golden*| mdo   | mdo     | mdo     | mdo  | mdo      |
|                               |               | *mdo_HChap1*|                |       |         |         |      |          |
| Prunus persica                | ppe           | ppe         | *prunus_persica*       | ppe   | ppe     | *pper*  | ppe  |  pper        |
| Prunus dulcis / P. amygdalus  | almond        | *almond*      | *prunus_dulcis*      |       | pdul    | pdul    | pdul | pdul     |
| Prunus avium                  | wild cherry   | *wildcherry*  | *prunus_avium*       |       | pavi    | pavi    | pavi | pavi     |
| Prunus armeniaca              | apricot       | *apricot*    |                       |      | parm    | parm    | parm | parm     |
| Prunus cerasifera             | cherry plum   | *cherryplum*  |                      |     |         | pcer    | pcer | pcer     |
| Pyrus                         | pear          | *pear*        |                      |     |         | pcox    | pcox | pcox     |
| Prunus sibirica               | Siberian apricot | *siberianapricot* |               |     |  psib | psib    | psib | psib     |


```{r}

group = "fruitTrees"

```


```{r, mdo, results='asis'}


params_list <- list(
  plantName = 'mdo'
  , plantNameOut = "apple"
  , plantDirOut = file.path('..', 'reports', group, "apple")
  , flag = 1
)

env <- new.env()
list2env(params_list, envir = env)

child_content <- knitr::knit_child("09_translate-child_strict.Rmd", envir = env, quiet = FALSE)
cat(child_content)



```


```{r, ppe, results='asis'}

params_list <- list(
  plantName = 'ppe'
  , plantNameOut = "peach"
  , plantDirOut = file.path('..', 'reports', group, "peach")
  , flag = 1
)

env <- new.env()
list2env(params_list, envir = env)

child_content <- knitr::knit_child("09_translate-child_strict.Rmd", envir = env, quiet = FALSE)
cat(child_content)


```

```{r, pdul, results='asis'}

params_list <- list(
  plantName = 'pdul'
  , plantNameOut = "almond"
  , plantDirOut = file.path('..', 'reports', group, "almond")
  , flag = 2
)

# note: in compara - geneID and prot ID are completely different

env <- new.env()
list2env(params_list, envir = env)

child_content <- knitr::knit_child("09_translate-child_strict.Rmd", envir = env, quiet = FALSE)
cat(child_content)


```


```{r, pavi, results='asis'}

params_list <- list(
  plantName = 'pavi'
  , plantNameOut = "wildcherry"
  , plantDirOut = file.path('..', 'reports', group, "wildcherry")
  , flag = 2
)

# note: in compara - geneID and prot ID are completely different

env <- new.env()
list2env(params_list, envir = env)

child_content <- knitr::knit_child("09_translate-child_strict.Rmd", envir = env, quiet = FALSE)
cat(child_content)


```



```{r, parm, results='asis'}


params_list <- list(
  plantName = 'parm'
  , plantNameOut = "apricot"
  , plantDirOut = file.path('..', 'reports', group, "apricot")
  , flag = 3
)

# note: in compara - geneID and prot ID are completely different

env <- new.env()
list2env(params_list, envir = env)

child_content <- knitr::knit_child("09_translate-child_strict.Rmd", envir = env, quiet = FALSE)
cat(child_content)


```




```{r, pcox, results='asis'}


params_list <- list(
  plantName = 'pcox'
  , plantNameOut = "pear"
  , plantDirOut = file.path('..', 'reports', group, "pear")
  , flag = 4
)


env <- new.env()
list2env(params_list, envir = env)

child_content <- knitr::knit_child("09_translate-child_strict.Rmd", envir = env, quiet = FALSE)
cat(child_content)


```


```{r, pcer, results='asis'}


params_list <- list(
  plantName = 'pcer'
  , plantNameOut = "cherryplum"
  , plantDirOut = file.path('..', 'reports', group, "cherryplum")
  , flag = 4
)

# note: in compara - geneID and prot ID are completely different

env <- new.env()
list2env(params_list, envir = env)

child_content <- knitr::knit_child("09_translate-child_strict.Rmd", envir = env, quiet = FALSE)
cat(child_content)


```


```{r, psib, results='asis'}


params_list <- list(
  plantName = 'psib'
  , plantNameOut = "siberianapricot"
  , plantDirOut = file.path('..', 'reports', group, "siberianapricot")
  , flag = 3
)


env <- new.env()
list2env(params_list, envir = env)

child_content <- knitr::knit_child("09_translate-child_strict.Rmd", envir = env, quiet = FALSE)
cat(child_content)


```






# SessionInfo

```{r,  echo=TRUE, warning=FALSE, message=FALSE}

sessionInfo()

```


